\scshape Evolutionary dynamics, equilibrium selection, and what population and quantitative genetics have taught evolutionary game theory

### 2022-10-02 manuscript.tex
added file and edits

%%
How to introduce maximization here? Hamilton wanted a maximization function. Lewontin spelled out clearly some of the important issues involved in maximization. JMS started his papers with the idea that game theory could provide solutions that didn't need group selection (or even initially kin selection) that involved individual-level maximization.
%%
Yes, start with the JMS justification for need of ESS and his dismal of Lewontin: this brings up the conflict between group/individual/kin and the issue of maximization and what to maximize.
%%
maybe I need to think more about my outline to get moving more quickly. I have a good start with mentioning ESS, early refs, and two primary ``confusions'': namely that ESS and the levels and/or kin/group issues are related and that ESS can't handle other evolutionary forces. I use the way that maynard smith frames the ESS papers to motivate Q1; for Q2, I can use feldman/nowak papers to motivate maybe?

the flow for the intro should be something like: maynard smith setup this false dichotomy: ESS is about ``individual'' level selection when really there are two issues, ESS as tool of optimality and then the application of that tool to kin/group models. this requires explaining how ESS is a tool of optimality and how a particular optimality measure, namely lineage growth rate, accounts for kin and group selection since it can captures selection at all levels from the gene on up. accounting for other evolutionary forces is more difficult but one way forward uses the fixation probability and allows mutation/demography etc to affect fixation probabilities and hence transition rates.

@@

Subsequent work by population geneticists revealed however that mean fitness can decrease due to frequency-dependent selection \cite{Wright:1955,Lewontin:1970} or recombination among multiple genetic loci \cite{Kojima:Kelleher:1961,Moran:1964,Karlin:1975}. Further, studies in the 1960s of the rates of molecular evolution \cite{Zuckerkandl:Pauling:1965,King:Jukes:1969} and levels of polymorphism \cite{Harris:1966,Lewontin:Hubby:1966} in a number of species spurred the development of the neutral \cite{Kimura:1968,Kimura:1983:book} and nearly neutral \cite{Ohta:1974,Ohta:1992} theories \cite{Ohta:Gillespie:1996} of molecular evolution. These theories posited that many mutations are weakly affected by natural selection, and thus their fate is governed mostly by genetic drift \cite{Kimura:1983:book}. Consequently, by the 1970s, evolutionary biologists were heavily debating both the relative role of natural selection versus other forces in shaping evolutionary patterns \cite{Gillespie:Langley:1974,Gillespie:1978}, and some biologists including Lewontin specifically questioned the importance of fitness maximization \cite{Karlin:1975,Gould:Lewontin:1979}. Maynard Smith was a clear proponent of fitness maximization and viewed it natural to ``[assume] that evolution has occurred by natural selection'' \cite[p. 31]{MaynardSmith:1978}. Thus, he proposed the ESS method developed by Price and himself as the appropriate tool for predicting phenotypic evolution and particularly for social traits. While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis (i.e., no dominance, epistasis, etc) and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are reasonable. This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate theory development in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}. As genomic data have become available for an increasing diverse set of species, some have argued that further advances in the evolution of behavior and other complex traits requires evolutionary theory that incorporates these data \cite[e.g.,][]{Springer:Crespi:2011,Rittschof:Robinson:2014,Akcay:Linksvayer:2015,Cunningham:2020}. Question \ref{q:II} arises here and asks whether the ESS method can be modified to accommodate other evolutionary forces such as mutation and recombination. I will argue below that the ESS method is more general than originally imagined by Maynard Smith and Price \cite{Maynard-Smith:Price:1973} in that in can in fact be incorporated into an evolutionary framework where natural selection and other evolutionary forces combine to shape the short and long-term evolution of traits.

@@

later asserted that since ``[i]t is assumed that evolution has occurred by natural selection'' \cite[p. 31]{MaynardSmith:1978},

the ESS method only identifies how natural selection shapes phenotypic traits to optimize fitness \cite[p. 42]{MaynardSmith:1978} and whereas population genetics can be applied to study ``selectively neutral'' molecular traits that are subject to mutation and genetic drift \cite[p. 37]{MaynardSmith:1978}.

This lead to / is similar to phenotypic gambit and idea of long-term evolution where genetics are inconsequential. Ah right but really maybe my argument is then (see below) that population genetic model that include many forces can be converted into long-term evolution models.

@@

In 1972, Price \cite{Price:1972:ftns} clarified the meaning the of the FTNS using his eponymous equation:

>>>>

%%
the flow for the intro should be something like: maynard smith setup this false dichotomy: ESS is about ``individual'' level selection when really there are two issues, ESS as tool of optimality and then the application of that tool to kin/group models. this requires explaining how ESS is a tool of optimality and how a particular optimality measure, namely lineage growth rate, accounts for kin and group selection since it can captures selection at all levels from the gene on up. accounting for other evolutionary forces is more difficult but one way forward uses the fixation probability and allows mutation/demography etc to affect fixation probabilities and hence transition rates.

@@

Subsequent work by population geneticists revealed however that mean fitness can decrease due to frequency-dependent selection \cite{Wright:1955,Lewontin:1970} or recombination among multiple genetic loci \cite{Kojima:Kelleher:1961,Moran:1964,Karlin:1975,Akin:1979}. Further, studies in the 1960s of the rates of molecular evolution \cite{Zuckerkandl:Pauling:1965,King:Jukes:1969} and levels of polymorphism \cite{Harris:1966,Lewontin:Hubby:1966} in a number of species spurred the development of the neutral \cite{Kimura:1968,Kimura:1983:book} and nearly neutral \cite{Ohta:1974,Ohta:1992} theories \cite{Ohta:Gillespie:1996} of molecular evolution. These theories posited that many mutations are weakly affected by natural selection, and thus their fate is governed mostly by genetic drift \cite{Kimura:1983:book}. Consequently, by the 1970s, evolutionary biologists were heavily debating both the relative role of natural selection versus other forces in shaping evolutionary patterns \cite{Gillespie:Langley:1974,Gillespie:1978}, and some biologists including Lewontin specifically questioned the importance of fitness maximization \cite{Karlin:1975,Gould:Lewontin:1979}. Maynard Smith was a clear proponent of fitness maximization and viewed it natural to ``[assume] that evolution has occurred by natural selection'' \cite[p. 31]{MaynardSmith:1978}. Thus, he proposed the ESS method developed by Price and himself as the appropriate tool for predicting phenotypic evolution and particularly for social traits. While Maynard Smith acknowledged that ESS models make a number of biological assumptions including that traits have a simple genetic basis (i.e., no dominance, epistasis, etc) and that appropriate genetic variation exits \cite{MaynardSmith:1978}, he believed that the limitations imposed by these assumptions are reasonable. This view that simplifying away genetic complexities or constraints is generally reasonable was termed the ``phenotypic gambit'' by Alan Grafen \cite{Grafen:1984} and came to dominate evolutionary theory in behavioral and evolutionary ecology \cite{Houston:McNamara:1999}. The near-singular focus of evolutionary game theory and the ESS method on natural selection has become less defensible given the recent and increasing abundance of genomic and transcriptomic data for a diverse set of species \cite[e.g.,][]{Kapheim:Pan:2015,Mikheyev:Linksvayer:2015,Warner:Mikheyev:2017,Kocher:Mallarino:2018,Warner:Mikheyev:2019}; specifically, some have argued that understanding these data requires moving beyond the gambit with more explicit consideration of complex genetic and demographic mechanisms \cite[e.g.,][]{Springer:Crespi:2011,Rittschof:Robinson:2014,Akcay:Linksvayer:2015,Cunningham:2020}. Question \ref{q:II} arises here and asks whether the ESS method can be modified to accommodate other evolutionary forces such as mutation and recombination. I will argue below that the ESS method is more general than originally imagined by Maynard Smith and Price \cite{Maynard-Smith:Price:1973} in that in can in fact be incorporated into an evolutionary framework where natural selection and other evolutionary forces combine to shape the short and long-term evolution of traits.

\section{Long-term evolution}

Following the development of the ESS approach by Maynard Smith and Price \cite{Maynard-Smith:Price:1973}, mathematical biologists evaluated the approach in light of the older and established theory population genetics. Population genetic models can explicitly address how both natural selection and genetic mechanisms including mutation, segregation, and recombination affect allele and genotype frequency dynamics over time. These models typically assume that evolution changes the frequencies of a fixed set of genotypes and the emergence of new genotypes through the large-effect mutations (i.e., the creation of new alleles or loci) occurs on a longer timescale than that of the model. For this reason, these population genetic models are often said to represent a ``short-term'' evolutionary process. Population genetic models showed that \cite{Cavalli-Sforza:Feldman:1978,Eshel:1982}

Long-term evolution cites \cite{Eshel:1996,Hammerstein:1996,Weissing:1996}

Though the idea of fitness optimization by itself has some intuitive appeal, high fitness individuals produce more surviving offspring who themselves are likely to be more fit than average, it needs an explicit evolutionary model in order to be evaluated mathematically and quantitatively.

@@

maybe my argument is then (see below) that population genetic model that include many forces can be converted into long-term evolution models.

@@
